Energy and Momentum of a Density-Driven Overflow in the Samoan Passage

### 2023-02-15 doc/northern_sill_energetics_r0.tex
test latexdiff

In this study we estimate the energy and momentum budget of flow across a major sill in the Samoan Passage using high horizontal resolution, towed, observations.
Results from a two-dimensional model are used to corroborate the analysis.
In the following, we give a short overview of the abyssal flow through the Samoan Passage and one of its major overflows (section~\ref{sec:ExperimentalDetails}\ref{sec:StudyRegion}), present towed observations of this overflow (section~\ref{sec:ExperimentalDetails}\ref{sec:Observations}),  and outline the setup of a two-dimensional numerical model simulating the dense overflow (section~\ref{sec:ExperimentalDetails}\ref{sec:Model}) to help interpret the observations.
After discussing the energy of the flow in terms of the Bernoulli equation (section~\ref{sec:Energetics}\ref{sec:Bernoulli}), a baroclinic energy equation is introduced (section~\ref{sec:Energetics}\ref{sec:EnergyEquation}).
Flow steadiness is discussed as a precursor to the following analysis in section~\ref{sec:Results}\ref{ssec:FlowSteadiness}.
Both energy frameworks are applied to the observed and modeled overflow to obtain energy budgets in section~\ref{sec:Results}\ref{ssec:EnergyBudgetResults}.
Form drag is calculated and evaluated against the energy budgets (\ref{sec:Results}\ref{ssec:FormDrag}).
Upward momentum flux estimates are presented in section~\ref{sec:Results}\ref{ssec:MomentumFluxes}.
The results are discussed and compared to observations of other high drag flows in section~\ref{sec:Discussion}.

@@

% Towyo Observations
\begin{figure*}
\centerline{\includegraphics[width=33pc]{towyo_overview.pdf}}
\caption{
Towyo-sections across the northern sill from 2012 (left) and 2014 (right).
({\bf a}) Potential temperature $\theta$ (color) and vertical velocity $w$ (black and white arrows showing upward/downward velocities, respectively, with scale given to lower right) with profile markers and a number of time stamps at top.
({\bf b}) Northward velocity $v$ (color) and $\sigma_4=45.94$\,kg\,m$^{-3}$ isopycnal tracing the upper interface (black contour).
({\bf c}) Vertical velocity $w$ (color) and $\sigma_4$ isopycnals at $10^{-2}$\,kg\,m$^{-3}$ intervals.
({\bf d}) Square of vertical shear $(\partial v/\partial z)^2$ (color) and isopycnals from panel c.
({\bf e}) Turbulent dissipation $\varepsilon$ from Thorpe-scale estimates.
({\bf f}) Barotropic tide prediction \citep[TPXO,][]{egberterofeeva02} for times and locations along the section.
Note the sharp transition in measured properties and tidal phase for the 2014 section at km\,12 where the instrument had to be recovered for a few hours.
}
\label{fig:TowyoObservations}
\end{figure*}

@@

\begin{figure*}
\centerline{\includegraphics[width=39pc]{model_snapshot.pdf}}
\caption{Model snapshot after 100 hours spinup time. Potential temperature is contoured in all four panels in 0.05$^{\circ}$C intervals starting at 0.7$^{\circ}$C. Thicker contours show the 0.8 and 0.9$^{\circ}$C isotherms.
(a) Potential temperature.
(b) Horizontal velocity.
(c) Vertical velocity.
(d) Turbulent dissipation $\varepsilon$ based on a parameterization acting on vertical density instabilities \citep[KL10,][]{klymaklegg10}.
}
\label{fig:ModelSnapshot}
\end{figure*}

@@

\subsection{Bernoulli flux}\label{sec:Bernoulli}
Treating the overflow as a single layer flow with the waters above at rest, we start out with the steady shallow water equations in one dimension:
\begin{equation}
    v \frac{\partial v}{\partial y} + g^\prime \frac{\partial}{\partial y} (\delta + h) = 0\ ,
    \label{eq:ShallowWater}
\end{equation}
where $\delta$ is the thickness of the layer, $v$ is the horizontal layer velocity along coordinate $y$, $h$ is the elevation of the topography, and $g^\prime=g \Delta \rho / \rho$ expresses the density difference $\Delta \rho$ between the bottom layer and waters aloft.
Neglecting entrainment leads to constant volume transport $Q$ of the dense bottom layer:
\begin{equation}
\frac{\partial (v \delta)}{\partial y} = \frac{\partial Q}{\partial y} = 0\ .
\end{equation}
Integrating \eqref{eq:ShallowWater} along the flow ($y$-coordinate) results in the Bernoulli function describing the sum of kinetic and potential energy of the system which is conserved for an isolated single layer except for dissipative regions like hydraulic jumps:
\begin{equation}
    B = \frac{v^2}{2} + g^\prime \delta + g^\prime h\ .
\end{equation}
The change in the energy flux associated with the transport of the Bernoulli function $F=Q B$, or Bernoulli flux, between upstream and downstream of a dissipative region over a flat bottom is
\begin{align}
\begin{split}
\Delta F &= Q_u B_u -Q_d B_d = Q(B_u - B_d) \\
    &= v_u \delta_u (\frac{v_u^2}{2} + g^\prime \delta_u - \frac{v_d^2}{2} -g^\prime \delta_d)\ ,
\end{split}
\label{eq:BernoulliNoEntrainment}
\end{align}
with subscripts $u$ and $d$ denoting upstream and downstream of a jump.
If entrainment is allowed then the volume flux changes and the drop is
\begin{equation}
\Delta F = v_u \delta_u (\frac{v_u^2}{2} + g^\prime_u \delta_u) - v_d \delta_d(\frac{v_d^2}{2} -g^\prime_d \delta_d)\ .
\label{eq:BernoulliEntrainment}
\end{equation}
Note that in \eqref{eq:BernoulliEntrainment} the bottom depth is the same between upstream and downstream.
We can express the energy drop including changes in bottom depth by adding the $h$ term:
\begin{equation}
\Delta F = \frac{v_u^3 \delta_u}{2} + v_u g^\prime_u (\delta_u^2 + h_u \delta_u) - \frac{v_d^3 \delta_d}{2} -v_d g^\prime_d (\delta_d^2 + h_d \delta_d)\ .
\label{eq:BernoulliEntrainmentVariableBottom}
\end{equation}
We can calculate $\Delta F$ following \eqref{eq:BernoulliEntrainmentVariableBottom} for various points upstream and downstream in model and observations, however, as we define a single layer $g^\prime$ and single layer velocity $v$, the results will be somewhat coarse.
As we will show in section~\ref{sec:Results}\ref{ssec:EnergyBudgetResults}, the drop in Bernoulli flux only converges to a meaningful result when sufficiently averaged in time.
% Furthermore, in this Bernoulli formulation of the overflow energetics we lack any information on how the energy loss occurs; whether it all dissipates within the overflow layer or whether some energy may radiate upwards in the form of an internal gravity wave energy flux.
We thus turn to a more detailed description of the overflow energetics in the next section.
Nevertheless, we expect results from these two approaches to be broadly comparable with each other.

@@

Vertical pressure work and internal wave fluxes, the small-scale subset of the pressure work term, are mostly upward and concentrated within the dense bottom layer.
The total pressure work term integrated along isopycnals between kilometer~0 and 17 differs quite substantially between observations and model (Fig.~\ref{fig:VerticalIWIntegrated}a), especially for the densest layers where it shows a downward flux of energy for the 2012 towyo. 
Pressure work within the dense layer varies between -4 and 4\,kW\,m$^{-1}$. 
Both towyo sections and the model show upward energy flux due to the pressure work term near the interface and diminishing magnitudes towards and beyond the interface.
The disagreement between the two towyos and the model makes the vertical pressure work term the least consistent term in the energy budget.
Integrated along isopycnals, the vertical energy flux due to smaller-scale internal waves is directed upwards and reaches between 1~and 2\,kW\,m$^{-1}$ within the bottom layer in the observations and somewhat smaller magnitudes in the model (Fig.~\ref{fig:VerticalIWIntegrated}b).
In both cases, vertical internal wave fluxes diminish close to zero past the upper interface of the overflow layer, indicating that the high frequency waves do not radiate much energy aloft outside the overflow layer.

@@

\begin{figure*}
\centerline{\includegraphics[width=27pc]{bottom_pressure.pdf}}
\caption{Pressure anomaly in observations and model.
a) Baroclinic pressure anomaly (colors) and isopycnals (contours) in the 2012 towyo transect.
The thick contour shows the $\sigma_4=45.94$\,kg\,m$^{-3}$ isopycnal previously defined as the upper layer interface.
The shaded area above the bottom shows depths not reached by the CTD observations and where constant density was assumed in the bottom pressure calculation.
b) Baroclinic bottom pressure in observations and model offset by constant factors (model:~60\,N\,m$^{-2}$, towyo~2012:~-130\,N\,m$^{-2}$, towyo~2014:~-150\,N\,m$^{-2}$) for visualization purposes.
This relatively large offset in absolute baroclinic bottom pressure between model and observations is due to model initialization with the observed density structure and partial draining of the initial stratification during model spinup as discussed in section~\ref{sec:ExperimentalDetails}\ref{sec:Model}.
Faint black lines show bottom pressure for each model time step of the analysis period.
The pink line shows the time mean model bottom pressure for the same period.
Blue and purple lines show bottom pressure in the observations. 
}
\label{fig:BottomPressure}
\end{figure*}

@@

\appendix[B]
\appendixtitle{Small-scale internal wave fluxes}
We work with locally defined velocity and pressure perturbations $\mathbf{u''}$ and $p''$ as detailed in section~\ref{sec:Energetics}\ref{sec:EnergyEquation} in \eqref{eq:BaroclinicPressureDecomposition} and \eqref{eq:BaroclinicoVelocityDecomposition} to investigate the role of smaller-scale internal waves for the energetics of the dense layer.
Varying the window size in the calculation of local mean profiles between 3 and 8\,km in the model changes the magnitude of the horizontally integrated vertical wave flux by about a factor of two with diminishing energy flux for smaller window sizes corresponding to smaller lateral scales.

@@

Diffusive background fluxes are explicitly set in the model through eddy viscosities $\nu_H$, $\nu_V$ and eddy diffusivities $\kappa_H$, $\kappa_V$ acting horizontally and vertically on momentum and mass, respectively.
These terms are inherently small as the bulk of the mixing is accomplished through the KL10 mixing parameterization (see section~\ref{sec:ExperimentalDetails}\ref{sec:Model}).
Even when calculated with the much larger turbulent diffusivities from the KL10 parameterization, which are calculated as $K_z = \Gamma \varepsilon N^{-2}$ based on a flux coefficient $\Gamma=0.2$ \citep{klymaklegg10}, the diffusive fluxes of baroclinic potential and kinetic energy are only $\mathcal{O}(1)$\,W\,m$^{-1}$ when integrated over the budget boundaries.
They are thus orders of magnitude smaller than other budget terms and therefore neglected.
Diffusive fluxes of baroclinic potential and kinetic energy in the observations can be estimated in a similar way and are of the same order of magnitude as in the model as the input parameters for the calculation are of similar size.
We note that using a larger flux coefficient, as may be necessary for near-bottom turbulence \citep[e.g.][]{ijichietal20,spingysetal21}, does not change the orders-of-magnitude difference to other terms in the energy budget.

>>>>

In this study we estimate the energy and momentum budget of flow across a major sill in the Samoan Passage using high resolution, towed, observations.
Results from a two-dimensional model are used to corroborate the analysis.
In the following, we give a short overview of the abyssal flow through the Samoan Passage and one of its major overflows (section~\ref{sec:ExperimentalDetails}\ref{sec:StudyRegion}), present towed observations of this overflow (section~\ref{sec:ExperimentalDetails}\ref{sec:Observations}),  and outline the setup of a two-dimensional numerical model simulating the dense overflow (section~\ref{sec:ExperimentalDetails}\ref{sec:Model}) to help interpret the observations.
After discussing the energy of the flow in terms of the Bernoulli equation (section~\ref{sec:Energetics}\ref{sec:Bernoulli}), a baroclinic energy equation is introduced (section~\ref{sec:Energetics}\ref{sec:EnergyEquation}).
Both frameworks are applied to the observed and modeled overflow for energy budgets in section~\ref{sec:Results}\ref{ssec:EnergyBudgetResults}.
Form drag is calculated and evaluated against the energy budgets (\ref{sec:Results}\ref{ssec:FormDrag}).
Upward momentum flux estimates are presented in section~\ref{sec:Results}\ref{ssec:MomentumFluxes}.
The results are discussed and compared to observations of other high drag flows in section~\ref{sec:Discussion}.

@@

% Towyo Observations
\begin{figure*}
\centerline{\includegraphics[width=39pc]{towyo_overview.pdf}}
\caption{
Towyo-sections across the northern sill from 2012 (left) and 2014 (right).
({\bf a}) Potential temperature $\theta$ (color) and vertical velocity $w$ (black and white arrows showing upward/downward velocities, respectively, with scale given to lower right) with profile markers and a number of time stamps at top.
({\bf b}) Northward velocity $v$ (color) and $\sigma_4=45.94$\,kg\,m$^{-3}$ isopycnal tracing the upper interface (black contour).
({\bf c}) Square of vertical shear $(\partial v/\partial z)^2$ (color) and isopycnal from panel b.
({\bf d}) Turbulent dissipation $\varepsilon$ from Thorpe-scale estimates.
({\bf e}) Barotropic tide prediction \citep[TPXO,][]{egberterofeeva02} for times and locations along the section.
Note the sharp transition in measured properties and tidal phase for the 2014 section at km\,12 where the instrument had to be recovered for a few hours.
}
\label{fig:TowyoObservations}
\end{figure*}

@@

\begin{figure*}
\centerline{\includegraphics[width=39pc]{model_snapshot.pdf}}
\caption{Model snapshot after 100 hours spinup time. Potential temperature is contoured in all four panels in 0.05$^{\circ}$C intervals starting at 0.7$^{\circ}$C. Thicker contours show the 0.8 and 0.9$^{\circ}$C isotherms.
(a) Potential temperature.
(b) Northward velocity.
(c) Vertical velocity.
(d) Turbulent dissipation $\varepsilon$ based on a parameterization acting on vertical density instabilities \citep[KL10,][]{klymaklegg10}.
}
\label{fig:ModelSnapshot}
\end{figure*}

@@

\subsection{Bernoulli flux}\label{sec:Bernoulli}
Treating the overflow as a single layer flow with the waters above at rest, we start out with the steady shallow water equations in one dimension:
\begin{equation}
    v \frac{\partial v}{\partial y} + g^\prime \frac{\partial}{\partial y} (\delta + h) = 0\ ,
    \label{eq:ShallowWater}
\end{equation}
where $\delta$ is the thickness of the layer, $v$ is the horizontal layer velocity along coordinate $y$, $h$ is the elevation of the topography, and $g^\prime=g \Delta \rho / \rho$ expresses the density difference $\Delta \rho$ between the bottom layer and waters aloft.
Neglecting entrainment leads to constant volume transport $Q$ of the dense bottom layer:
\begin{equation}
\frac{\partial (v \delta)}{\partial y} = \frac{\partial Q}{\partial y} = 0\ .
\end{equation}
Integrating \eqref{eq:ShallowWater} along the flow ($y$-coordinate) results in the Bernoulli function describing the sum of kinetic and potential energy of the system which is conserved for an isolated single layer except for dissipative regions like hydraulic jumps:
\begin{equation}
    B = \frac{v^2}{2} + g^\prime \delta + g^\prime h\ .
\end{equation}
The change in the energy flux associated with the transport of the Bernoulli function $F=Q B$, or Bernoulli flux, between upstream and downstream of a dissipative region over a flat bottom is
\begin{align}
\begin{split}
\Delta F &= Q_u B_u -Q_d B_d = Q(B_u - B_d) \\
    &= v_u d_u (\frac{v_u^2}{2} + g^\prime \delta_u - \frac{v_d^2}{2} -g^\prime \delta_d)\ ,
\end{split}
\label{eq:BernoulliNoEntrainment}
\end{align}
with subscripts $u$ and $d$ denoting upstream and downstream of a jump.
If entrainment is allowed then the volume flux changes and the drop is
\begin{equation}
\Delta F = v_u \delta_u (\frac{v_u^2}{2} + g^\prime_u \delta_u) - v_d \delta_d(\frac{v_d^2}{2} -g^\prime_d \delta_d)\ .
\label{eq:BernoulliEntrainment}
\end{equation}
Note that in \eqref{eq:BernoulliEntrainment} the bottom depth is the same between upstream and downstream.
We can express the energy drop including changes in bottom depth by adding the $h$ term:
\begin{equation}
\Delta F = \frac{v_u^3 \delta_u}{2} + v_u g^\prime_u (\delta_u^2 + h_u \delta_u) - \frac{v_d^3 \delta_d}{2} -v_d g^\prime_d (\delta_d^2 + h_d \delta_d)\ .
\label{eq:BernoulliEntrainmentVariableBottom}
\end{equation}
We can calculate $\Delta F$ following \eqref{eq:BernoulliEntrainmentVariableBottom} for various points upstream and downstream in model and observations, however, as we define a single layer $g^\prime$ and single layer velocity $v$, the results will be somewhat coarse.
As we will show in section~\ref{sec:Results}\ref{ssec:EnergyBudgetResults}, the drop in Bernoulli flux only converges to a meaningful result when sufficiently averaged in time.
% Furthermore, in this Bernoulli formulation of the overflow energetics we lack any information on how the energy loss occurs; whether it all dissipates within the overflow layer or whether some energy may radiate upwards in the form of an internal gravity wave energy flux.
We thus turn to a more detailed description of the overflow energetics in the next section.
Nevertheless, we expect results from these two approaches to be broadly comparable with each other.

@@

Vertical pressure work and internal wave fluxes, the small-scale subset of the pressure work term, are mostly upward and concentrated within the dense bottom layer.
The total pressure work term integrated along isopycnals between kilometer~0 and 17 differs quite substantially between observations and model (Fig.~\ref{fig:VerticalIWIntegrated}a), especially for the densest layers where it shows a downward flux of energy for the 2012 towyo. 
Fluxes within the dense layer vary between -4 and 4\,kW\,m$^{-1}$. 
Both towyo sections and the model show upward energy flux due to the pressure work term near the interface and diminishing magnitudes towards and beyond the interface.
The disagreement between the two towyos and the model makes the vertical pressure work term the least consistent term in the energy budget.
Integrated along isopycnals, the vertical energy flux due to smaller-scale internal waves is directed upwards and reaches between 1~and 2\,kW\,m$^{-1}$ within the bottom layer in the observations and somewhat smaller magnitudes in the model (Fig.~\ref{fig:VerticalIWIntegrated}b).
In both cases, vertical internal wave fluxes diminish close to zero past the upper interface of the overflow layer, indicating that the high frequency waves do not radiate much energy aloft outside the overflow layer.

@@

\begin{figure*}
\centerline{\includegraphics[width=27pc]{bottom_pressure.pdf}}
\caption{Pressure anomaly in observations and model.
a) Baroclinic pressure anomaly (colors) and isopycnals (contours) in the 2012 towyo transect.
The thick contour shows the $\sigma_4=45.94$\,kg\,m$^{-3}$ isopycnal previously defined as the upper layer interface.
The shaded area above the bottom shows depths not reached by the CTD observations and where constant density was assumed in the bottom pressure calculation.
b) Baroclinic bottom pressure in observations and model offset by constant factors for visualization purposes.
Faint black lines show bottom pressure for each model time step of the analysis period.
The pink line shows the time mean model bottom pressure for the same period.
Blue and purple lines show bottom pressure in the observations. 
}
\label{fig:BottomPressure}
\end{figure*}

@@

\appendix[B]
\appendixtitle{Small-scale internal wave fluxes}
We work with locally defined velocity and pressure perturbations $\mathbf{u''}$ and $p''$ as detailed in section~\ref{sec:Energetics}\ref{sec:EnergyEquation} in \eqref{eq:BaroclinicPressureDecomposition} and \eqref{eq:BaroclinicoVelocityDecomposition} to investigate the role of smaller-scale internal waves for the energetics of the dense layer.
Varying the window size in the calculation of local mean profiles between 3 and 8\,km in the model changes the magnitude of the horizontally integrated vertical wave flux as shown in Fig.~\ref{fig:VerticalIWIntegrated}b by about a factor of two with diminishing energy flux for smaller window sizes corresponding to smaller lateral scales.

@@

Diffusive background fluxes are explicitly set in the model through eddy viscosities $\nu_H$, $\nu_V$ and eddy diffusivities $\kappa_H$, $\kappa_V$ acting horizontally and vertically on momentum and mass, respectively.
These terms are small and therefore neglected in the budget.
Estimates of diffusive fluxes in the observations are also small and neglected.
